---
title: "Project 1"
subtitle: "Noam Ghenassia, Maïlis Fontani, Julia Morin"
output:
  html_document: default
  pdf_document: default
date: "2022-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In this project, we will analyse data from the IPUMS USA database, which provides access to over sixty integrated, high-precision samples of the American population drawn from sixteen federal censuses (Source : <https://usa.ipums.org/usa-action/variables/group>)

We wanted in particular to observe the effects of education and origin on the income in the American population. We were inspired by a study "The Employment of Low-Skilled Immigrant Men in the United States" (Brian Duncan Stephen J. Trejo, American Economic Review, May 2012) that claims that among people with the lowest education levels, immigrants exhibit substantially higher rates of employment than comparable natives.

In this purpose, we selected a sample from year 2000. It consists of 2808457 observations together with the origin of the individuals (they are classified in 9 categories: white, black/African american, Chinese...), their income and their educational levels (12 categories from *No Schooling* to *5+ years of college*).

## Analysis

To find relations between our different variables, we start to compare the income and the origin of the individuals. Let us start with a boxplot of the income for each of the origin to see how income is distributed. Here we plot the logarithm of the incomes : this prevents the outliers with higher income from hiding the variability among the lower levels of income, and allows for all the orders of magnitude of income to fit in the same plot.

```{r boxplot race income, echo = FALSE, message=FALSE}
library(ggplot2)
library(tidyverse)
library(dplyr)

data <- read.csv("usa_00006.csv")

subsamp <- select(data, RACE, INCTOT, EDUC)

subsamp["RACE"][subsamp["RACE"] == 1] <- "White"
subsamp["RACE"][subsamp["RACE"] == 2] <- "Black/African American"
subsamp["RACE"][subsamp["RACE"] == 3] <- "American Indian/Alaska Native"
subsamp["RACE"][subsamp["RACE"] == 4] <- "Chinese"
subsamp["RACE"][subsamp["RACE"] == 5] <- "Japanese"
subsamp["RACE"][subsamp["RACE"] == 6] <- "Other Asian/Pacific Islander"
subsamp["RACE"][subsamp["RACE"] == 7] <- "Other race, nec	"
subsamp["RACE"][subsamp["RACE"] == 8] <- "More than 2 races"
subsamp["RACE"][subsamp["RACE"] == 9] <- "More than 2 races"

subsamp = subsamp[subsamp$INCTOT >= 1, ]
subsamp = subsamp[subsamp$INCTOT != 9999999, ]

Origin <- subsamp$RACE
Income <- log(subsamp$INCTOT)

ggplot( mapping = aes(x = Origin, y = Income)) +
  geom_boxplot() + theme(axis.text=element_text(size=5)) + labs(y= "log Income in dollars/year ", x = "Origin")

```

The diagram does not seem to show outstanding differences between the different origins. The highest income median is for the Japanese population (with an income around exp(10) = 22000 US-dollars). The three following highest median are for the Chinese, White and other Asians or Pacific Islanders. It can also be observed that the highest variance stands for the Chinese. Japanese and Chinese are also the populations with less outliers with a high income. The lowest incomes of the Japanese population are also way higher than those of the other categories.

We now process to the same diagram with two boxplots showing the income for two categories of our observations : individuals that attempted school until the 12th grade and the ones that did at least one year of college.

```{r boxplot education income,  echo = FALSE,message=FALSE}

subsamp <- select(data, RACE, INCTOT, EDUC)

  subsamp["EDUC"][subsamp["EDUC"] == 00] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 01] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 02] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 03] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 04] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 05] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 06] <- "Grade 12 or less"
  subsamp["EDUC"][subsamp["EDUC"] == 07] <- "1+ year of college"
  subsamp["EDUC"][subsamp["EDUC"] == 08] <- "1+ year of college"
  subsamp["EDUC"][subsamp["EDUC"] == 09] <- "1+ year of college"
  subsamp["EDUC"][subsamp["EDUC"] == 10] <- "1+ year of college"
  subsamp["EDUC"][subsamp["EDUC"] == 11] <- "1+ year of college"
  
  
  subsamp = subsamp[subsamp$INCTOT >= 1, ]
  subsamp = subsamp[subsamp$INCTOT != 9999999, ]
  
  Income <- log(subsamp$INCTOT)
  
  subsamp$EDUC <- factor(subsamp$EDUC, levels=c("Grade 12 or less","1+ year of college"))
  ggplot( mapping = aes(x = subsamp$EDUC, y = Income), cex.axis=0.2) +
    geom_boxplot() + labs(y= "log Income in dollars/year ", x = "Education attainment")
```

The second category, i.e. the workers that had a better level of education gets highest incomes overall. Variance and outliers seems to be similar. To have a more precise idea, we can do more boxplots with a broadest range of category to be more precise on the flow of income :

```{r income vs origin boxplot, echo = FALSE, message=FALSE}
library(ggplot2)
library(tidyverse)
library(dplyr)

subsamp <- select(data, RACE, INCTOT, EDUC)
subsamp["EDUC"][subsamp["EDUC"] == 00] <- "N/A or no schooling"
subsamp["EDUC"][subsamp["EDUC"] == 01] <- "Nursery school to grade 4"
subsamp["EDUC"][subsamp["EDUC"] == 02] <- "Grade 5, 6, 7, or 8"
subsamp["EDUC"][subsamp["EDUC"] == 03] <- "Grade 9"
subsamp["EDUC"][subsamp["EDUC"] == 04] <- "Grade 10"
subsamp["EDUC"][subsamp["EDUC"] == 05] <- "Grade 11"
subsamp["EDUC"][subsamp["EDUC"] == 06] <- "Grade 12"
subsamp["EDUC"][subsamp["EDUC"] == 07] <- "1 year of college"
subsamp["EDUC"][subsamp["EDUC"] == 08] <- "2 years of college"
subsamp["EDUC"][subsamp["EDUC"] == 09] <- "3 years of college"
subsamp["EDUC"][subsamp["EDUC"] == 10] <- "4 years of college"
subsamp["EDUC"][subsamp["EDUC"] == 11] <- "5+ years of college"

subsamp = subsamp[subsamp$INCTOT >= 1, ]
subsamp = subsamp[subsamp$INCTOT != 9999999, ]


subsamp$EDUC <- factor(subsamp$EDUC, levels=c("N/A or no schooling", "Nursery school to grade 4", "Grade 5, 6, 7, or 8","Grade 9","Grade 10","Grade 11","Grade 12","1 year of college", "2 years of college", "3 years of college", "4 years of college","5+ years of college"))
ggplot( mapping = aes(x = subsamp$EDUC, y =log(subsamp$INCTOT)), cex.axis=0.2) +
  geom_boxplot() + labs(y= "log Income in dollars/year", x = "Education attainment")





```

Here we can observe that the main gap occurs, as expected, after finishing high school. The income thenseems to steadily increase with each year after grade 12. On the other hand, the median income appears to decrease for people who started, but did not finish high school, compared with people who did not attend high school at all.

To get a more holistic vision of our data, we produced a heatmap that links all the variables in one diagram.

```{r presentation of the datas, echo = FALSE, message=FALSE}
library(heatmaply)
library(ggplot2)
library(tidyverse)
library(dplyr)

subsamp <- select(data, RACE, INCTOT, EDUC)

subsamp["RACE"][subsamp["RACE"] == 1] <- "White"
subsamp["RACE"][subsamp["RACE"] == 2] <- "Black/African American"
subsamp["RACE"][subsamp["RACE"] == 3] <- "American Indian/ Alaska Native"
subsamp["RACE"][subsamp["RACE"] == 4] <- "Chinese"
subsamp["RACE"][subsamp["RACE"] == 5] <- "Japanese"
subsamp["RACE"][subsamp["RACE"] == 6] <- "Other Asian/Pacific Islander"
subsamp["RACE"][subsamp["RACE"] == 7] <- "Other race, nec	"
subsamp["RACE"][subsamp["RACE"] == 8] <- "More than 2 races"
subsamp["RACE"][subsamp["RACE"] == 9] <- "More than 2 races"

subsamp["EDUC"][subsamp["EDUC"] == 00] <- "N/A or no schooling"
subsamp["EDUC"][subsamp["EDUC"] == 01] <- "Nursery school to grade 4"
subsamp["EDUC"][subsamp["EDUC"] == 02] <- "Grade 5, 6, 7, or 8"
subsamp["EDUC"][subsamp["EDUC"] == 03] <- "Grade 9"
subsamp["EDUC"][subsamp["EDUC"] == 04] <- "Grade 10"
subsamp["EDUC"][subsamp["EDUC"] == 05] <- "Grade 11"
subsamp["EDUC"][subsamp["EDUC"] == 06] <- "Grade 12"
subsamp["EDUC"][subsamp["EDUC"] == 07] <- "1 year of college"
subsamp["EDUC"][subsamp["EDUC"] == 08] <- "2 years of college"
subsamp["EDUC"][subsamp["EDUC"] == 09] <- "3 years of college"
subsamp["EDUC"][subsamp["EDUC"] == 10] <- "4 years of college"
subsamp["EDUC"][subsamp["EDUC"] == 11] <- "5+ years of college"

subsamp = subsamp[subsamp$INCTOT >= 1, ]
subsamp = subsamp[subsamp$INCTOT != 9999999, ]
subsamp$EDUC <- factor(subsamp$EDUC, levels=c("N/A or no schooling",
                                              "Nursery school to grade 4",
                                              "Grade 5, 6, 7, or 8",
                                              "Grade 9","Grade 10",
                                              "Grade 11",
                                              "Grade 12",
                                              "1 year of college",
                                              "2 years of college",
                                              "3 years of college",
                                              "4 years of college",
                                              "5+ years of college"))

mat <- aggregate(x=as.numeric(subsamp$INCTOT),
                 by=list(subsamp$RACE,subsamp$EDUC),
                 FUN=mean)
colnames(mat) <- c("Origin", "Education", "Average_income")


ggplot(mat, aes(x = Origin, y = Education, fill = log(Average_income))) +
  theme(axis.text=element_text(size=5)) + geom_tile()
x11(width = 20, height = 4)

```

We observe that the difference of income is more important between the rows than the columns. This is linked to the analysis we had with the boxplots. The incomes tends to depend more one the education levels than the origins.

We will now perform Kernel Density Estimation for the income distribution, for each education level or origin, and will compare these plots. We use a Gaussian kernel.

```{r KDE income education, echo = FALSE, message=FALSE}
education <- data$EDUC
income <- data$INCTOT
race <- data$RACE

data_income_educ <- data.frame(education, income)

data_income_educ <- data_income_educ[data_income_educ$income != 9999999, ] 
data_income_educ <- data_income_educ[data_income_educ$income != 0, ]

bw = 1600
max= 100000

df <- data.frame()

data_1 <- filter(data_income_educ, education == "1")
Y1 <- density(as.numeric(data_1$income), bw=bw,
              n=10000, from=0, to=max)
df1 <- data.frame(income = matrix(unlist(Y1[1])), density1 =  matrix(unlist(Y1[2])))
df <- df1

data_2 <- filter(data_income_educ, education == "2")
Y2 <- density(as.numeric(data_2$income), bw=bw,
              n=10000, from=0, to=max)
df2 <- data.frame(density2 =  matrix(unlist(Y2[2])))
df <-cbind(df, df2)

data_3 <- filter(data_income_educ, education == "3")
Y3 <- density(as.numeric(data_3$income), bw=bw,
              n=10000, from=0, to=max)
df3 <- data.frame(density3 =  matrix(unlist(Y3[2])))
df <-cbind(df, df3)

data_4 <- filter(data_income_educ, education == "4")
Y4 <- density(as.numeric(data_4$income), bw=bw,
              n=10000, from=0, to=max)
df4 <- data.frame(density4 =  matrix(unlist(Y4[2])))
df <-cbind(df, df4)

data_5 <- filter(data_income_educ, education == "5")
Y5 <- density(as.numeric(data_5$income), bw=bw,
              n=10000, from=0, to=max)
df5 <- data.frame(density5 =  matrix(unlist(Y5[2])))
df <-cbind(df, df5)

data_6 <- filter(data_income_educ, education == "6")
Y6 <- density(as.numeric(data_6$income), bw=bw,
              n=10000, from=0, to=max)
df6 <- data.frame(density6 =  matrix(unlist(Y6[2])))
df <-cbind(df, df6)

data_7 <- filter(data_income_educ, education == "7")
Y7 <- density(as.numeric(data_7$income), bw=bw,
              n=10000, from=0, to=max)
df7 <- data.frame(density7 =  matrix(unlist(Y7[2])))
df <-cbind(df, df7)

data_8 <- filter(data_income_educ, education == "8")
Y8 <- density(as.numeric(data_8$income), bw=bw,
              n=10000, from=0, to=max)
df8 <- data.frame(density8 =  matrix(unlist(Y8[2])))
df <-cbind(df, df8)

data_10 <- filter(data_income_educ, education == "10")
Y10 <- density(as.numeric(data_10$income), bw=bw,
              n=10000, from=0, to=max)
df10 <- data.frame(density10 =  matrix(unlist(Y10[2])))
df <-cbind(df, df10)

data_11 <- filter(data_income_educ, education == "11")
Y11 <- density(as.numeric(data_11$income), bw=bw,
              n=10000, from=0, to=max)
df11 <- data.frame(density11 =  matrix(unlist(Y11[2])))
df <-cbind(df, df11)



ggplot(data=df, mappping=aes(x=income)) + labs(y= "Density ", x = "Income") +
  geom_line(aes(x=income, y=density1, colour="Nursery school to grade 4")) +
  geom_line(aes(x=income, y=density2, colour="Grade 5, 6, 7, or 8")) +
  geom_line(aes(x=income, y=density3, colour="Grade 9")) +
  geom_line(aes(x=income, y=density4, colour="Grade 10")) +
  geom_line(aes(x=income, y=density5, colour="Grade 11")) +
  geom_line(aes(x=income, y=density6, colour="Grade 12")) +
  geom_line(aes(x=income, y=density7, colour="1 year of college")) +
  geom_line(aes(x=income, y=density8, colour="2 years of college")) +
  geom_line(aes(x=income, y=density10, colour="4 years of college")) +
  geom_line(aes(x=income, y=density11, colour="5+ years of college")) +
  scale_colour_manual("", 
                      breaks = c("Nursery school to grade 4",
                                 "Grade 5, 6, 7, or 8",
                                 "Grade 9", "Grade 10",
                                 "Grade 11",
                                 "Grade 12",
                                 "1 year of college",
                                 "2 years of college",
                                 "3 years of college",
                                 "4 years of college",
                                 "5+ years of college"),
                      values = c("red", "green", "blue", "yellow", "orange",
                                 "pink", "purple", "brown", "black", "magenta")) 
```

The bandwidth for the KDE was chosen by simply trying several values. The retained value of 1600 seems to offer a good tradeoff between bias and variance : indeed, the curves are relatively smooth, but still show a small peak on multiples of 10000. This is likely due to the fact that it is more common to earn multiples of 10000 dollars, or that people rounded up their answer. Therefore, we can observe that while being relatively smooth, the curves are still sensitive enough to show relevant information about the data. An alternative way of choosing a bandwidth would be to simply perform cross validation on this specific parameter.

Clearly the five first curves (representing the KDE of the incomes for individuals that stopped school after the 12th grade) are high for low income and decrease very quickly. On the other hand, there seems to be a clear change of behaviour for the curves of subjects that finished high school : the curves of the Grade 12 and 1 year of college categories have their peak around the same value as those of lower education levels, but are a lot more right skewed. For even higher levels of education, the trend is the same : the maximum density is achieved for increasingly large income values, but more importantly, we can observe that the curves become more spreaded, indicating a greater variability among the same categories.

Here is the diagram for the KDE of the density of the incomes with respect to the different origins.

```{r KDE income origin, echo = FALSE, message=FALSE}

bw = 1600
max= 100000
education <- data$EDUC
income <- data$INCTOT
race <- data$RACE
data_income_race <- data.frame(race, income)
df <- data.frame()

data_1 <- filter(data_income_race, race == "1")
Y1 <- density(as.numeric(data_1$income), bw=bw,
              n=10000, from=0, to=max)
df1 <- data.frame(income = matrix(unlist(Y1[1])), density1 =  matrix(unlist(Y1[2])))
df <- df1

data_2 <- filter(data_income_race, race == "2")
Y2 <- density(as.numeric(data_2$income), bw=bw,
              n=10000, from=0, to=max)
df2 <- data.frame(density2 =  matrix(unlist(Y2[2])))
df <-cbind(df, df2)

data_3 <- filter(data_income_race, race == "3")
Y3 <- density(as.numeric(data_3$income), bw=bw,
              n=10000, from=0, to=max)
df3 <- data.frame(density3 =  matrix(unlist(Y3[2])))
df <-cbind(df, df3)

data_4 <- filter(data_income_race, race == "4")
Y4 <- density(as.numeric(data_4$income), bw=bw,
              n=10000, from=0, to=max)
df4 <- data.frame(density4 =  matrix(unlist(Y4[2])))
df <-cbind(df, df4)

data_5 <- filter(data_income_race, race == "5")
Y5 <- density(as.numeric(data_5$income), bw=bw,
              n=10000, from=0, to=max)
df5 <- data.frame(density5 =  matrix(unlist(Y5[2])))
df <-cbind(df, df5)

data_6 <- filter(data_income_race, race == "6")
Y6 <- density(as.numeric(data_6$income), bw=bw,
              n=10000, from=0, to=max)
df6 <- data.frame(density6 =  matrix(unlist(Y6[2])))
df <-cbind(df, df6)

data_7 <- filter(data_income_race, race == "7")
Y7 <- density(as.numeric(data_7$income), bw=bw,
              n=10000, from=0, to=max)
df7 <- data.frame(density7 =  matrix(unlist(Y7[2])))
df <-cbind(df, df7)

data_8 <- filter(data_income_race, race == "8")
Y8 <- density(as.numeric(data_8$income), bw=bw,
              n=10000, from=0, to=max)
df8 <- data.frame(density8 =  matrix(unlist(Y8[2])))
df <-cbind(df, df8)

data_9 <- filter(data_income_race, race == "9")
Y9 <- density(as.numeric(data_9$income), bw=bw,
              n=10000, from=0, to=max)
df9 <- data.frame(density9 =  matrix(unlist(Y9[2])))
df <-cbind(df, df9)


ggplot(data=df, mappping=aes(x=income)) + labs(y= "Density ", x = "Income") +
  geom_line(aes(x=income, y=density1, colour="White")) +
  geom_line(aes(x=income, y=density2, colour="Black/African American")) +
  geom_line(aes(x=income, y=density3, colour="American Indian or Alaska Native")) +
  geom_line(aes(x=income, y=density4, colour="Chinese")) +
  geom_line(aes(x=income, y=density5, colour= "Japanese")) +
  geom_line(aes(x=income, y=density6, colour="Other Asian or Pacific Islander")) +
  geom_line(aes(x=income, y=density7, colour="Other race, nec")) +
  geom_line(aes(x=income, y=density8, colour="Two major races")) +
  geom_line(aes(x=income, y=density9, colour="Three or more races")) +
  scale_colour_manual("", 
                      breaks = c("White",
                                 "Black/African American",
                                 "American Indian or Alaska Native", "Chinese",
                                 "Japanese",
                                 "Other Asian or Pacific Islander",
                                 "Other race, nec",
                                 "Two major races",
                                 "Three or more races"),
                      values = c("red", "green", "blue", "yellow", "orange",
                                 "pink", "purple", "brown", "black")) 
```

Here the density estimations tend to be much more similar to each other. In fact there are again the curves standing for the Chinese, Japanese and other Asian or Pacific Islander that show a higher density for higher incomes. However, the difference is negligible compared with the variability observed among the education levels.

## **Conclusion**

This is not enough to comment on the claim of the study we presented at the beginning. In fact, we should more concentrate on the rate of employment than the income to face the problematic the study had. We also could take a subset of data concerning the part of the population with less education (until grade 10th approximately, as we saw with KDE) and analyse this subsets with respect to the different origins.

Overall, we observed a clear effect of the education level on the income. On the other hand, the origin of the subject appeared to be almost irrelevant to explain the variability in income. To explain the remaining variability between the different origins, it would be interesting to observe how each origin category is distributed among the education levels.
